---
title: "Building credential chains"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building credential chains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(azr)
```

A *credential chain* is an ordered list of authentication methods. When you ask
for a token, `azr` walks the chain from top to bottom and returns the first
credential that successfully authenticates. This lets the same code run
unchanged on a laptop (interactive login), in CI (a client secret), and on a
managed Azure host (managed identity).

This vignette shows how to build and use chains.

## The default chain

Most of the time you don't build a chain at all — the high-level functions use
`default_credential_chain()` automatically:

```{r}
default_credential_chain()
```

Each line is one *entry*: a name on the left and the credential it will try on
the right. The chain is attempted top to bottom, so non-interactive methods
(client secret, workload identity, managed identity, Azure CLI) come before the
interactive browser/device-code flows.

You get a token without ever mentioning the chain:

```{r}
token <- get_token(scope = "https://graph.microsoft.com/.default")
```

`get_token()`, `get_request_authorizer()`, `get_token_provider()`,
`get_credential_auth()`, and `get_credential_provider()` all take a `chain =`
argument that defaults to `default_credential_chain()`.

## Building a custom chain

Use `credential_chain()` to build your own. Pass **named** entries; each entry
is either:

* a credential **class** (e.g. `ClientSecretCredential`),
* an already-constructed `Credential` **instance**.

```{r}
chain <- credential_chain(
  client_secret = ClientSecretCredential,
  azure_cli     = AzureCLICredential
)

chain
```

The names (`client_secret`, `azure_cli`) are labels -- they show up when the
chain prints and in verbose/error output, so pick something recognisable.

Constructing a chain does **no** authentication. Credential configuration is
validated when a chain entry is tried, just before the first token request:

```{r}
token <- get_token(
  scope = "https://graph.microsoft.com/.default",
  chain = chain
)
```

## How context flows into each entry

When you call `get_token(scope = ..., tenant_id = ..., ...)`, those values form a
*context* that is handed to each class entry as it is tried. Each credential only
receives the context arguments its own constructor accepts -- a
`ManagedIdentityCredential` takes `scope` and `client_id`, so a `tenant_id` in
the context is simply ignored for that entry.

This is why a single `get_token(scope = ...)` call works across every method in
the chain: the scope is forwarded to whichever credential ends up succeeding.

Constructed instances are different: they are used **as-is**. Use an instance
when a chain entry has its own configuration:

```{r}
prod_secret <- ClientSecretCredential$new(
  tenant_id = "11111111-1111-1111-1111-111111111111",
  client_id = "app-registration-id",
  client_secret = "super-secret-value"
)

chain <- credential_chain(
  prod_secret = prod_secret,
  azure_cli = AzureCLICredential
)

chain
```

Credential objects validate their configuration at construction, so an
incomplete object fails fast rather than at token time:

```{r}
ClientSecretCredential$new(client_secret = NULL)
```

Note that sensitive values are redacted when a credential prints:

```{r}
ClientSecretCredential$new(client_secret = "super-secret-value")
```

## Using a pre-built instance

```{r}
my_cli <- AzureCLICredential$new(auto_login = FALSE)

chain <- credential_chain(
  configured_cli = my_cli,
  fallback       = DeviceCodeCredential
)
```

## Reusing a chain with `DefaultCredential`

`DefaultCredential` wraps a chain in an object with a lazily-resolved provider.
The chain is resolved once on first use and the winning credential is reused for
later calls:

```{r}
cred <- DefaultCredential$new(
  scope = "https://graph.microsoft.com/.default",
  chain = chain
)

# First call resolves the chain; later calls reuse the same provider.
token <- cred$get_token()
resp  <- httr2::req_perform(cred$req_auth(httr2::request(url)))
```

## Seeing which credential wins

Set the `chain_verbose` option (or the `AZR_CHAIN_VERBOSE` environment variable)
to watch the chain being walked — each entry tried, which fail, and which one
finally returns a token:

```{r}
options(azr.chain_verbose = TRUE)
get_token(scope = "https://graph.microsoft.com/.default", chain = chain)
```

If every entry fails, `get_token()` aborts with a combined report listing each
named entry and the error it raised, so you can see why the whole chain came up
empty.
