---
title: "Declaring Azure Storage datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Declaring Azure Storage datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(azr)
```

A *dataset* is a named, environment-aware pointer to data in Azure Storage. You
declare **where** the data lives once, and then ask `azr` to build the concrete
URI for whichever environment (tier) you are running in. A *catalog* groups
related datasets so they can be stored, shared, and resolved together.

This separates two concerns that usually get tangled together in pipeline code:

* the **stable identity** of a dataset (its name, container, path, format), and
* the **environment-specific** storage account it currently lives in.

## A single dataset

`az_dataset()` declares one dataset. Only the storage account varies by tier —
the container, path, scheme, and format are shared across every environment:

```{r}
orders <- az_dataset(
  name      = "orders",
  scheme    = "abfss",
  container = "raw",
  storage   = list(prod = "stprod001", preprod = "stpreprod001"),
  path      = "sales/orders",
  format    = "delta"
)

orders
```

`name` must match `^[a-z][a-z0-9_]*$` so it can be used as a stable lookup key,
`path` must be non-empty (a dataset cannot point at a container root), and
`format` must be one of `delta`, `parquet`, `csv`, `tsv`, `json`, `avro`, `orc`,
or `text`.

## Building a URI

`az_dataset_uri()` turns a dataset into a concrete URI for one tier. The default
`uri_type = "hadoop"` produces the ABFS form used by Spark, Flink, Trino, and any
`hadoop-azure` consumer:

```{r}
az_dataset_uri(orders, tier = "prod")
#> "abfss://raw@stprod001.dfs.core.windows.net/sales/orders"
```

Switch tier to point the same dataset at a different storage account, or ask for
the `https` form:

```{r}
az_dataset_uri(orders, tier = "preprod", uri_type = "https")
#> "https://stpreprod001.dfs.core.windows.net/raw/sales/orders"
```

Requesting a tier the dataset doesn't declare is an error that lists the tiers it
does have, so misconfiguration fails loudly rather than silently reading the
wrong environment.

### The default tier

`tier` defaults to the `dataset_tier` option, so production code rarely needs to
pass it. Set it once via `options(azr.dataset_tier = ...)` or the
`AZR_DATASET_TIER` environment variable (default `"prod"`) — for example, a
pre-production job can export `AZR_DATASET_TIER=preprod` and every
`az_dataset_uri()` call follows.

## Resolving to a manifest

A URI alone isn't enough to *load* data — a reader also needs the format.
`az_resolve_dataset()` returns an `az_dataset_manifest` carrying the `name`,
`uri`, and `format` together (call `as.list()` for a plain R list):

```{r}
m <- az_resolve_dataset(orders, tier = "prod")
m
#> <az_dataset_manifest:orders>
#> uri: abfss://raw@stprod001.dfs.core.windows.net/sales/orders
#> format: delta
```

This is exactly the shape a reader like `sparklyr::spark_read_source()` wants:

```{r}
m <- az_resolve_dataset(orders, tier = "prod")
spark_read_source(sc, name = m@name, source = m@format, path = m@uri)
```

## Creating a dataset from a URI

If you already have a full storage URI, `az_dataset_from_uri()` parses it into a
dataset. The parsed storage account is bound to `tier`, and the `format` is
inferred from the URI's file extension (or `_delta_log` segment) when you don't
pass it:

```{r}
file <- az_dataset_from_uri(
  "abfss://raw@stprod001.dfs.core.windows.net/sales/orders/part.parquet",
  name = "orders_file"
)

file
#> <az_dataset:orders_file>
#> scheme: abfss
#> container: raw
#> path: sales/orders/part.parquet
#> format: parquet
#> storage: prod=stprod001
```

A directory URI is ambiguous (its format can't be inferred), so pass `format`
explicitly in that case:

```{r}
az_dataset_from_uri(
  "abfss://raw@stprod001.dfs.core.windows.net/sales/orders",
  name   = "orders",
  format = "delta"
)
```

To make the dataset multi-tier, supply additional accounts via `storage`:

```{r}
az_dataset_from_uri(
  "abfss://raw@stprod001.dfs.core.windows.net/sales/orders",
  name    = "orders",
  format  = "delta",
  tier    = "prod",
  storage = list(preprod = "stpreprod001")
)
```

## Grouping datasets in a catalog

An `az_catalog()` holds an ordered collection of datasets with unique names. It
behaves like a named container: index with `[[`, and use `names()` and
`length()`:

```{r}
catalog <- az_catalog(datasets = list(orders, file))

names(catalog)
#> "orders" "orders_file"
length(catalog)
#> 2
catalog[["orders"]]
```

`az_dataset_uri()` and `az_resolve_dataset()` both dispatch on a catalog. With no
`name`, they return one entry per dataset, keyed by name:

```{r}
az_dataset_uri(catalog, tier = "prod")
#> orders      "abfss://raw@stprod001.dfs.core.windows.net/sales/orders"
#> orders_file "abfss://raw@stprod001.dfs.core.windows.net/sales/orders/part.parquet"
```

Pass `name` to resolve a single dataset out of the catalog:

```{r}
az_dataset_uri(catalog, tier = "prod", name = "orders")
az_resolve_dataset(catalog, tier = "prod", name = "orders")
```

## Storing a catalog as JSON

A catalog round-trips to JSON, so you can keep dataset definitions in version
control and load them at runtime. `az_catalog_write()` serialises a catalog and
`az_catalog_read()` reads it back:

```{r}
az_catalog_write(catalog, "datasets.json")

catalog <- az_catalog_read("datasets.json")
```

The JSON shape is one object per dataset under a top-level `datasets` array:

```json
{
  "datasets": [
    {
      "name": "orders",
      "scheme": "abfss",
      "container": "raw",
      "storage": { "prod": "stprod001", "preprod": "stpreprod001" },
      "path": "sales/orders",
      "format": "delta"
    }
  ]
}
```

`az_catalog_read()` validates every entry, reporting which dataset (by index and
name) is invalid and why — so a malformed catalog fails at load time rather than
when a pipeline later tries to read a dataset.

## Putting it together

A typical flow: keep a `datasets.json` in the repo, read it into a catalog, and
resolve the dataset you need for the current tier.

```{r}
catalog <- az_catalog_read("datasets.json")

m <- az_resolve_dataset(catalog, name = "orders")  # tier from AZR_DATASET_TIER
df <- spark_read_source(sc, name = m@name, source = m@format, path = m@uri)
```

The same code runs unchanged in every environment; only the `dataset_tier`
option (or `AZR_DATASET_TIER`) decides which storage account the URIs point at.
