---
title: "Configuring azr options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configuring azr options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(azr)
```

`azr` reads a small set of package options that control verbosity and default
behaviour. Each one can be set three different ways, checked in this order:

1. an in-session value set via R's `options()`,
2. an environment variable, or
3. a built-in default.

## Available options

| Name | R option | Env variable | Default | Controls |
|------|----------|--------------|---------|----------|
| `chain_verbose` | `azr.chain_verbose` | `AZR_CHAIN_VERBOSE` | `FALSE` | Verbose credential-chain discovery (see the credential chains vignette) |
| `api_verbose` | `azr.api_verbose` | `AZR_API_VERBOSE` | `FALSE` | `>>>`/`<<<` request and response logging in `api_client` |
| `cli_auto_login` | `azr.cli_auto_login` | `AZR_CLI_AUTO_LOGIN` | `FALSE` | Whether `AzureCLICredential` runs `az login` automatically |
| `dataset_tier` | `azr.dataset_tier` | `AZR_DATASET_TIER` | `"prod"` | Default tier for `azr_dataset_uri()` / `azr_resolve_dataset()` (see the datasets vignette) |

## Setting an option

Use base R's `options()` with the `azr.` prefix:

```{r}
options(azr.chain_verbose = TRUE)
```

or set the corresponding environment variable before starting R (handy for CI
and containers, since it doesn't require any R code):

```{sh}
export AZR_DATASET_TIER=preprod
```

A value set via `options()` takes priority over the environment variable, which
in turn takes priority over the built-in default.

## Inspecting the current configuration

`azr_options()` prints every option, its resolved value, where that value came
from (`[default]`, `[envvar]`, or `[option]`), and the underlying environment
variable:

```{r}
azr_options()
#> ── azr options ──────────────────────────────────────
#> api_verbose = "FALSE" (default)
#>   `AZR_API_VERBOSE`: (not set)
#> chain_verbose = "FALSE" (default)
#>   `AZR_CHAIN_VERBOSE`: (not set)
#> cli_auto_login = "FALSE" (default)
#>   `AZR_CLI_AUTO_LOGIN`: (not set)
#> dataset_tier = "prod" (default)
#>   `AZR_DATASET_TIER`: (not set)
```

After setting `options(azr.dataset_tier = "preprod")`, the same call shows where
the new value came from:

```{r}
options(azr.dataset_tier = "preprod")
azr_options()
#> dataset_tier = "preprod" [option]
#>   `AZR_DATASET_TIER`: (not set)
```

`azr_options()` also returns this information invisibly as a `data.frame`, for
use in scripts or diagnostics:

```{r}
df <- azr_options()
df
#>           option   value  source            env_var env_value default
#> 1    api_verbose   FALSE default    AZR_API_VERBOSE      <NA>   FALSE
#> 2  chain_verbose   FALSE default  AZR_CHAIN_VERBOSE      <NA>   FALSE
#> 3 cli_auto_login   FALSE default AZR_CLI_AUTO_LOGIN      <NA>   FALSE
#> 4   dataset_tier preprod  option   AZR_DATASET_TIER      <NA>    prod
```

If any option held a sensitive value, `azr_options()` would redact it as
`"<hidden>"`; pass `mask = FALSE` to see raw values.

## Reading an option from code

Internally (and in your own code, if you build on `azr`), an option's resolved
value is read with `opts$get()`, following the same
value → `options()` → env var → default precedence:

```{r}
azr:::opts$get("dataset_tier")
#> "preprod"
```
